---
title: "CIRCRES/2025/326808R1"
output: html_notebook
---

```{r setup, include=TRUE, message = FALSE}

knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = TRUE
)



#BiocManager::install("GO.db", force = TRUE)
#BiocManager::install("chromVAR")

#library(chromVAR)

library(scCustomize)
library(BiocManager)
library(S4Vectors)
library(IRanges)
library(GenomicRanges)
library(AnnotationHub)
library(monocle3)
library(dplyr)
library(Seurat)
library(monocle3)
library(sctransform)
library(future)
library(EnsDb.Hsapiens.v86)
library(Matrix)
library(irlba)
library(Signac)
library(glmGamPoi)
library(batchelor)
library(AnnotationHub)
library(multtest)
library(metap)
library(Signac)
library(Seurat)
library(JASPAR2020)
library(TFBSTools)
library(patchwork)
library(BSgenome.Hsapiens.UCSC.hg38)
library(chromVAR)
# for plotting
library(ggplot2)
library(patchwork)
set.seed(1234)
library(tibble)
library(ggpubr)

library(future)
options(future.globals.maxSize = 32000 * 1024^2) # for 16 Gb RAM
set.seed(1234)

```

#Load in Data and create matrix

```{r}
data_dir.1 <- '../ATACseq SSBP PBMC/S11D2_results_MULTIOME/filtered_feature_bc_matrix'
data_dir.2 <- '../ATACseq SSBP PBMC/S11D3_results_MULTIOME/filtered_feature_bc_matrix'
data_dir.3 <- '../ATACseq SSBP PBMC/S13D2_results_MULTIOME/filtered_feature_bc_matrix'
data_dir.4 <- '../ATACseq SSBP PBMC/S13D3_results_MULTIOME/filtered_feature_bc_matrix'

fragpath.1 <- "../ATACseq SSBP PBMC/S11D2_atac_fragments.tsv.gz"
fragpath.2 <- "../ATACseq SSBP PBMC/S11D3_atac_fragments.tsv.gz"
fragpath.3 <- "../ATACseq SSBP PBMC/S13D2_atac_fragments.tsv.gz"
fragpath.4 <- "../ATACseq SSBP PBMC/S13D3_atac_fragments.tsv.gz"

list.files(data_dir.1)
expression.matrix.1 <- Read10X(data.dir = data_dir.1)
expression.matrix.2 <- Read10X(data.dir = data_dir.2)
expression.matrix.3 <- Read10X(data.dir = data_dir.3)
expression.matrix.4 <- Read10X(data.dir = data_dir.4)

data_dir.1
```

#RNA analysis

```{r message = FALSE}
# Extract RNA data
rna.counts.1 <- expression.matrix.1$'Gene Expression'
rna.counts.2 <- expression.matrix.2$'Gene Expression'
rna.counts.3 <- expression.matrix.3$'Gene Expression'
rna.counts.4 <- expression.matrix.4$'Gene Expression'

rna.1 <- CreateSeuratObject(counts = rna.counts.1, project = "SS-SL", assay = "RNA")
rna.2 <- CreateSeuratObject(counts = rna.counts.2, project = "SS-SD", assay = "RNA")
rna.3 <- CreateSeuratObject(counts = rna.counts.3, project = "SR-SL", assay = "RNA")
rna.4 <- CreateSeuratObject(counts = rna.counts.4, project = "SR-SD", assay = "RNA")

#rna.1[["percent.mt"]] <- PercentageFeatureSet(rna.1, pattern = "^MT-")
#rna.2[["percent.mt"]] <- PercentageFeatureSet(rna.2, pattern = "^MT-")
#rna.3[["percent.mt"]] <- PercentageFeatureSet(rna.3, pattern = "^MT-")
#rna.4[["percent.mt"]] <- PercentageFeatureSet(rna.4, pattern = "^MT-")
rna.1$dataset <- "SS-SL"
rna.2$dataset <- "SS-SD"
rna.3$dataset <- "SR-SL"
rna.4$dataset <- "SR-SD"
```

#Create ATAC objects for each sample

```{r}
atac.counts.1 <- expression.matrix.1$Peaks
atac.counts.2 <- expression.matrix.2$Peaks
atac.counts.3 <- expression.matrix.3$Peaks
atac.counts.4 <- expression.matrix.4$Peaks

#Convert to genomic ranges
grange.counts.1 <- StringToGRanges(regions = rownames(atac.counts.1), sep =c(":", "-"))
grange.counts.2 <- StringToGRanges(regions = rownames(atac.counts.2), sep =c(":", "-"))
grange.counts.3 <- StringToGRanges(regions = rownames(atac.counts.3), sep =c(":", "-"))
grange.counts.4 <- StringToGRanges(regions = rownames(atac.counts.4), sep =c(":", "-"))

#filter granges for standard chromosomes
standard_peaks <- seqnames(grange.counts.1) %in% standardChromosomes(grange.counts.1)
standard_peaks <- seqnames(grange.counts.2) %in% standardChromosomes(grange.counts.2)
standard_peaks <- seqnames(grange.counts.3) %in% standardChromosomes(grange.counts.3)
standard_peaks <- seqnames(grange.counts.4) %in% standardChromosomes(grange.counts.4)

grange.counts.1 <- grange.counts.1[standard_peaks]
grange.counts.2 <- grange.counts.2[standard_peaks]
grange.counts.3 <- grange.counts.3[standard_peaks]
grange.counts.4 <- grange.counts.4[standard_peaks]

#Create a unified set of peaks to quantify in each dataset
combined.peaks <- reduce(c(grange.counts.1, grange.counts.2, grange.counts.3, grange.counts.4))

#filter out bad peaks based on length
peakwidths <- width(combined.peaks)
combined.peaks <- combined.peaks[peakwidths < 10000 & peakwidths > 20]
combined.peaks


#Create Fragment objects
cells.1 <- colnames(expression.matrix.1$Peaks)
cells.2 <- colnames(expression.matrix.2$Peaks)
cells.3 <- colnames(expression.matrix.3$Peaks)
cells.4 <- colnames(expression.matrix.4$Peaks)

frag.1 <- CreateFragmentObject(
  path = fragpath.1,
  cells = cells.1
)
frag.2 <- CreateFragmentObject(
  path = fragpath.2,
  cells = cells.2
)
frag.3 <- CreateFragmentObject(
  path = fragpath.3,
  cells = cells.3
)
frag.4 <- CreateFragmentObject(
  path = fragpath.4,
  cells = cells.4
)

#New counts based on unions
new.atac.counts.1 <- FeatureMatrix(fragments = frag.1, features = combined.peaks, cells = cells.1)
new.atac.counts.2 <- FeatureMatrix(fragments = frag.2, features = combined.peaks, cells = cells.2)
new.atac.counts.3 <- FeatureMatrix(fragments = frag.3, features = combined.peaks, cells = cells.3)
new.atac.counts.4 <- FeatureMatrix(fragments = frag.4, features = combined.peaks, cells = cells.4)

rna.1[["ATAC"]] <- CreateChromatinAssay(
  counts = new.atac.counts.1, 
  sep = c(":", "-"), 
  fragments = fragpath.1, 
  min.cells = 10,
  genome = "hg38"
)

rna.2[["ATAC"]] <- CreateChromatinAssay(
  counts = new.atac.counts.2, 
  sep = c(":", "-"), 
  fragments = fragpath.2, 
  min.cells = 10,
  genome = "hg38"
)

rna.3[["ATAC"]] <- CreateChromatinAssay(
  counts = new.atac.counts.3, 
  sep = c(":", "-"), 
  fragments = fragpath.3, 
  min.cells = 10,
  genome = "hg38"
)

rna.4[["ATAC"]] <- CreateChromatinAssay(
  counts = new.atac.counts.4, 
  sep = c(":", "-"), 
  fragments = fragpath.4, 
  min.cells = 10,
  genome = "hg38"
)

rna.1[["percent.mt"]]<- PercentageFeatureSet(rna.1, pattern = "^MT-")
rna.2[["percent.mt"]]<- PercentageFeatureSet(rna.2, pattern = "^MT-")
rna.3[["percent.mt"]]<- PercentageFeatureSet(rna.3, pattern = "^MT-")
rna.4[["percent.mt"]]<- PercentageFeatureSet(rna.4, pattern = "^MT-")

pbmc.1 <- rna.1
pbmc.2 <- rna.2
pbmc.3 <- rna.3
pbmc.4 <- rna.4

granges(pbmc.1[["ATAC"]])
granges(pbmc.2[["ATAC"]])
granges(pbmc.3[["ATAC"]])
granges(pbmc.4[["ATAC"]])

#let's annotate some sequences
ah <- AnnotationHub()
query(ah, "EnsDb.Hsapiens.v98")
ensdb_v98 <- ah[["AH75011"]]
#extract gene annotations from EnsDb
annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98) 
#change to UCSC style
seqlevels(annotations) <- paste0('chr', seqlevels(annotations))
genome(annotations) <- "hg38"

seqlevelsStyle(pbmc.1[["ATAC"]]) <- "UCSC"


Annotation(pbmc.1[["ATAC"]]) <- annotations
Annotation(pbmc.2[["ATAC"]]) <- annotations
Annotation(pbmc.3[["ATAC"]]) <- annotations
Annotation(pbmc.4[["ATAC"]]) <- annotations

seqlevels(pbmc.1[["ATAC"]])

# Get GRanges object
gr1 <- granges(pbmc.1[["ATAC"]])
gr2 <- granges(pbmc.2[["ATAC"]])
gr3 <- granges(pbmc.3[["ATAC"]])
gr4 <- granges(pbmc.4[["ATAC"]])

# Identify peaks in standard chromosomes
keep_peaks.1 <- seqnames(gr1) %in% standardChromosomes(gr1)
keep_peaks.2 <- seqnames(gr2) %in% standardChromosomes(gr2)
keep_peaks.3 <- seqnames(gr3) %in% standardChromosomes(gr3)
keep_peaks.4 <- seqnames(gr4) %in% standardChromosomes(gr4)

# Subset the GRanges objects directly
gr1_filtered <- gr1[keep_peaks.1]
gr2_filtered <- gr2[keep_peaks.2]
gr3_filtered <- gr3[keep_peaks.3]
gr4_filtered <- gr4[keep_peaks.4]

# Assign back the filtered GRanges to ATAC assay
pbmc.1[["ATAC"]] <- subset(pbmc.1[["ATAC"]], features = names(gr1_filtered))
pbmc.2[["ATAC"]] <- subset(pbmc.2[["ATAC"]], features = names(gr2_filtered))
pbmc.3[["ATAC"]] <- subset(pbmc.3[["ATAC"]], features = names(gr3_filtered))
pbmc.4[["ATAC"]] <- subset(pbmc.4[["ATAC"]], features = names(gr4_filtered))

# Check seqlevels again
seqlevels(pbmc.1[["ATAC"]])

#get standard chromosomes
std_chroms <- paste0("chr", c(1:22, "X", "Y"))

# Get GRanges from ATAC assay
gr1 <- granges(pbmc.1[["ATAC"]])
gr2 <- granges(pbmc.2[["ATAC"]])
gr3 <- granges(pbmc.3[["ATAC"]])
gr4 <- granges(pbmc.4[["ATAC"]])

# Filter only standard chromosomes
gr1_filtered <- gr1[seqnames(gr1) %in% std_chroms]
gr2_filtered <- gr2[seqnames(gr2) %in% std_chroms]
gr3_filtered <- gr3[seqnames(gr3) %in% std_chroms]
gr4_filtered <- gr4[seqnames(gr4) %in% std_chroms]

seqlevels(gr1_filtered) <- seqlevelsInUse(gr1_filtered)
seqlevels(gr2_filtered) <- seqlevelsInUse(gr2_filtered)
seqlevels(gr3_filtered) <- seqlevelsInUse(gr3_filtered)
seqlevels(gr4_filtered) <- seqlevelsInUse(gr4_filtered)


# Verify the update
seqlevels(pbmc.1[["ATAC"]])
seqlevels(pbmc.2[["ATAC"]])
seqlevels(pbmc.3[["ATAC"]])
seqlevels(pbmc.4[["ATAC"]])

saveRDS(pbmc.1, "pbmc1_unanalyzed.rds")
saveRDS(pbmc.2, "pbmc2_unanalyzed.rds")
saveRDS(pbmc.3, "pbmc3_unanalyzed.rds")
saveRDS(pbmc.4, "pbmc4_unanalyzed.rds")

pbmc.1 <- readRDS("../ATACseq SSBP PBMC/pbmc1_unanalyzed.rds")
pbmc.2 <- readRDS("../ATACseq SSBP PBMC/pbmc2_unanalyzed.rds")
pbmc.3 <- readRDS("../ATACseq SSBP PBMC/pbmc3_unanalyzed.rds")
pbmc.4 <- readRDS("../ATACseq SSBP PBMC/pbmc4_unanalyzed.rds")
```

#QC and ATAC/RNA Analysis

```{r}
DefaultAssay(pbmc.1) <- "ATAC"
DefaultAssay(pbmc.2) <- "ATAC"
DefaultAssay(pbmc.3) <- "ATAC"
DefaultAssay(pbmc.4) <- "ATAC"

pbmc.1 <- NucleosomeSignal(pbmc.1)
pbmc.2 <- NucleosomeSignal(pbmc.2)
pbmc.3 <- NucleosomeSignal(pbmc.3)
pbmc.4 <- NucleosomeSignal(pbmc.4)

pbmc.1 <- TSSEnrichment(pbmc.1, fast = FALSE)
pbmc.2 <- TSSEnrichment(pbmc.2, fast = FALSE)
pbmc.3 <- TSSEnrichment(pbmc.3, fast = FALSE)
pbmc.4 <- TSSEnrichment(pbmc.4, fast = FALSE)

DensityScatter(pbmc.1, x = 'nCount_ATAC', y = 'TSS.enrichment', log_x = TRUE, quantiles = TRUE)
DensityScatter(pbmc.2, x = 'nCount_ATAC', y = 'TSS.enrichment', log_x = TRUE, quantiles = TRUE)
DensityScatter(pbmc.3, x = 'nCount_ATAC', y = 'TSS.enrichment', log_x = TRUE, quantiles = TRUE)
DensityScatter(pbmc.4, x = 'nCount_ATAC', y = 'TSS.enrichment', log_x = TRUE, quantiles = TRUE)

VlnPlot(pbmc.1, features = c("nCount_ATAC", "nCount_RNA", "percent.mt", "TSS.enrichment", "nucleosome_signal"), ncol = 5, log = TRUE, pt.size = 0) + NoLegend()
VlnPlot(pbmc.2, features = c("nCount_ATAC", "nCount_RNA", "percent.mt", "TSS.enrichment", "nucleosome_signal"), ncol = 5, log = TRUE, pt.size = 0) + NoLegend()
VlnPlot(pbmc.3, features = c("nCount_ATAC", "nCount_RNA", "percent.mt", "TSS.enrichment", "nucleosome_signal"), ncol = 5, log = TRUE, pt.size = 0) + NoLegend()
VlnPlot(pbmc.4, features = c("nCount_ATAC", "nCount_RNA", "percent.mt", "TSS.enrichment", "nucleosome_signal"), ncol = 5, log = TRUE, pt.size = 0) + NoLegend()

#filter out low quality cells
pbmc.1 <- subset(
  x = pbmc.1,
  subset = nCount_ATAC < 10000 &
    nCount_ATAC > 100 &
    nCount_RNA < 12000 &
    nCount_RNA > 500 &
    percent.mt < 15 & 
    nucleosome_signal < 2 &
    TSS.enrichment > 3
)
pbmc.2 <- subset(
  x = pbmc.2,
  subset = nCount_ATAC < 10000 &
    nCount_ATAC > 100 &
    nCount_RNA < 12000 &
    nCount_RNA > 500 &
    percent.mt < 15 & 
    nucleosome_signal < 2 &
    TSS.enrichment > 3
)
pbmc.3 <- subset(
  x = pbmc.3,
  subset = nCount_ATAC < 10000 &
    nCount_ATAC > 100 &
    nCount_RNA < 12000 &
    nCount_RNA > 500 &
    percent.mt < 15 & 
    nucleosome_signal < 2 &
    TSS.enrichment > 3
)
pbmc.4 <- subset(
  x = pbmc.4,
  subset = nCount_ATAC < 10000 &
    nCount_ATAC > 100 &
    nCount_RNA < 12000 &
    nCount_RNA > 500 &
    percent.mt < 15 & 
    nucleosome_signal < 2 &
    TSS.enrichment > 3
)


```

#Gene Expression Data Processing, merging objects, re-processing the merged object

```{r}
#normalize gene expression using SCTransform and reduce dimensionality using PCA
DefaultAssay(pbmc.1) <- "RNA"
DefaultAssay(pbmc.2) <- "RNA"
DefaultAssay(pbmc.3) <- "RNA"
DefaultAssay(pbmc.4) <- "RNA"

pbmc.1 <- NormalizeData(pbmc.1) %>% FindVariableFeatures() %>% ScaleData() 
pbmc.2 <- NormalizeData(pbmc.2) %>% FindVariableFeatures() %>% ScaleData() 
pbmc.3 <- NormalizeData(pbmc.3) %>% FindVariableFeatures() %>% ScaleData() 
pbmc.4 <- NormalizeData(pbmc.4) %>% FindVariableFeatures() %>% ScaleData() 

pbmc.1 <- RunPCA(pbmc.1) %>% RunUMAP(dims = 1:50, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')
pbmc.2 <- RunPCA(pbmc.2) %>% RunUMAP(dims = 1:50, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')
pbmc.3 <- RunPCA(pbmc.3) %>% RunUMAP(dims = 1:50, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')
pbmc.4 <- RunPCA(pbmc.4) %>% RunUMAP(dims = 1:50, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')



DefaultAssay(pbmc.1) <- "ATAC"
DefaultAssay(pbmc.2) <- "ATAC"
DefaultAssay(pbmc.3) <- "ATAC"
DefaultAssay(pbmc.4) <- "ATAC"

pbmc.1 <-  RunTFIDF(pbmc.1) %>% FindTopFeatures(min.cutoff = 5) %>% RunSVD() %>% RunUMAP (reduction = 'lsi', dims = 2:50, reduction.name = 'umap.atac', reduction.key = 'atacUMAP_')
pbmc.2 <-  RunTFIDF(pbmc.2) %>% FindTopFeatures(min.cutoff = 5) %>% RunSVD() %>% RunUMAP (reduction = 'lsi', dims = 2:50, reduction.name = 'umap.atac', reduction.key = 'atacUMAP_')
pbmc.3 <-  RunTFIDF(pbmc.3) %>% FindTopFeatures(min.cutoff = 5) %>% RunSVD() %>% RunUMAP (reduction = 'lsi', dims = 2:50, reduction.name = 'umap.atac', reduction.key = 'atacUMAP_')
pbmc.4 <-  RunTFIDF(pbmc.4) %>% FindTopFeatures(min.cutoff = 5) %>% RunSVD() %>% RunUMAP (reduction = 'lsi', dims = 2:50, reduction.name = 'umap.atac', reduction.key = 'atacUMAP_')

DefaultAssay(pbmc.1) <- "RNA"
DefaultAssay(pbmc.2) <- "RNA"
DefaultAssay(pbmc.3) <- "RNA"
DefaultAssay(pbmc.4) <- "RNA"

saveRDS(pbmc.1, "pbmc1_analyzed.rds")
saveRDS(pbmc.2, "pbmc2_analyzed.rds")
saveRDS(pbmc.3, "pbmc3_analyzed.rds")
saveRDS(pbmc.4, "pbmc4_analyzed.rds")


pbmc.1 <- readRDS("../ATACseq SSBP PBMC/pbmc1_analyzed.rds")
pbmc.2 <- readRDS("../ATACseq SSBP PBMC/pbmc2_analyzed.rds")
pbmc.3 <- readRDS("../ATACseq SSBP PBMC/pbmc3_analyzed.rds")
pbmc.4 <- readRDS("../ATACseq SSBP PBMC/pbmc4_analyzed.rds")

seqlevels(pbmc.1[["ATAC"]])

pbmc.combined <- merge(
  x = pbmc.1, 
  y = list(pbmc.2, pbmc.3, pbmc.4), 
  add.cell.ids = c("SS-SL", "SS-SD", "SR-SL", "SR-SD"), 
  project = "SSBP")

pbmc.combined <- NormalizeData(pbmc.combined) %>% FindVariableFeatures() %>% ScaleData() 
pbmc.combined <- RunPCA(pbmc.combined) %>% RunUMAP(dims = 1:50, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')

DefaultAssay(pbmc.combined) <- "ATAC"

pbmc.combined <-  RunTFIDF(pbmc.combined) %>% FindTopFeatures(min.cutoff = 5) %>% RunSVD() %>% RunUMAP (reduction = 'lsi', dims = 2:50, reduction.name = 'umap.atac', reduction.key = 'atacUMAP_')

```

#Integration RNA

```{r, fig.asp=.5, fig.width = '400px', fig.height='200px'}
library(harmony)
DefaultAssay(pbmc.combined) <- "RNA"
pbmc.combined <- RunHarmony(pbmc.combined, group.by.vars = "orig.ident", dims.use = 1:50, max.iter.harmony = 50)
pbmc.combined <- RunUMAP(pbmc.combined, reduction = "harmony", dims = 1:50)
pbmc.combined <- FindNeighbors(pbmc.combined, reduction = "harmony", dims = 1:50) %>% FindClusters(resolution = 0.6)

saveRDS(pbmc.combined, "rna_harmony_integrated_pbmc_combined.rds")
pbmc.combined <- readRDS("../ATACseq SSBP PBMC/rna_harmony_integrated_pbmc_combined.rds")
p1 <- UMAPPlot(pbmc.combined, group.by = "orig.ident")
p2 <- UMAPPlot(pbmc.combined, label = T)
p1+p2
```

#ATAC Non-linear dimension reduction with UMAP Visualization

```{r, fig.asp=1, fig.width = '200px'}
DefaultAssay(pbmc.combined) <- "ATAC"

p1 <- ElbowPlot(pbmc.combined, ndims = 50, reduction = "lsi")
p2 <- DepthCor(pbmc.combined, n = 50)
p1+p2

p3 <- DimPlot(pbmc.combined, group.by = "orig.ident", reduction = "umap") 
p4 <- FeaturePlot(pbmc.combined, reduction = "umap", c("LYZ", "NKG7", "CCL5", "IL32", "FCGR3A", "CD1C"))
p3
p4
```

#Integrate ATAC assay

```{r, fig.asp=1, fig.width = '200px'}
library(harmony)
pbmc.combined <- RunHarmony(pbmc.combined,
                            group.by.vars = "orig.ident",
                            assay.use = "ATAC",
                            reduction = "lsi",
                            dims.use = 2:50,
                            reduction.save = "harmony_atac",
                            project.dim = FALSE)
pbmc.combined <- RunUMAP(pbmc.combined, 
                         reduction = "harmony_atac",
                         dims = 1:ncol(Embeddings(pbmc.combined, "harmony_atac")),
                         reduction.name = "umap_harmony_atac",
                         reduction.key = "UMAPharmonyATAC_")

  p1 <- DimPlot(pbmc.combined,
                group.by = "orig.ident",
                reduction = "umap_harmony_atac") + ggtitle("ATAC Harmony UMAP")
  p2 <- FeaturePlot(pbmc.combined,
                    c("LYZ", "NKG7", "CCL5", "IL32", "FCGR3A", "CD1C"),
                    reduction = "umap_harmony_atac") & NoAxes() & NoLegend()
  p1 / p2
```

#Azimuth Annotation

```{r, fig.asp=.75, fig.width = '800px'}
#Install Azithmuth - web-tool that maps user-uploaded datasets
devtools::install_github("satijalab/azimuth", "seurat5")
library(Azimuth)
library(SeuratData)
#downgrade seurat to get ref and my object to line up
remotes::install_version(package = 'Seurat', version = package_version('5.3.0'))
library(Seurat)
packageVersion('Seurat')


InstallData("pbmcref")

#RunAzimuth function can take a Seurat object as input, pbmcref included in the function
DefaultAssay(pbmc.combined) <- "RNA"
pbmc.combined <- JoinLayers(pbmc.combined)
pbmc.combined <- RunAzimuth(pbmc.combined, reference = "pbmcref")

#Visualize Aximuth outputs
umap_cols <- DiscretePalette_scCustomize(num_colors = 26, palette = "glasbey", shuffle_pal = TRUE)

p1 <- UMAPPlot(pbmc.combined, group.by = "predicted.celltype.l2", label = TRUE, label.size = 4, repel = TRUE, cols = umap_cols) + theme(legend.text = element_text(size = 9, colour = "black"), legend.key.size = unit(5, "pt")) + guides( color = guide_legend(ncol = 1), color = guide_legend(override.aes = list(size = 10))) + ggtitle("SSBP PBMC: scATACseq")
p2 <- UMAPPlot(pbmc.combined, group.by = "orig.ident")
p1+p2
p1
#Add idents based on prediction
Idents(pbmc.combined) <- "predicted.celltype.l2"

p1 <- FeaturePlot(pbmc.combined, features = "CCR7", reduction = "umap")
p2 <- FeaturePlot(pbmc.combined, features = "FCGR3A", reduction = "umap")
p3 <- FeaturePlot(pbmc.combined, features = "ITGAX", reduction = "umap")
p4 <- FeaturePlot(pbmc.combined, features = "predictionscorecelltypel2_Treg", reduction = "umap")

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
DefaultAssay(pbmc.combined) <- "RNA"
DimPlot(pbmc.combined, label = TRUE, label.size = 3, reduction = "umap", cols = umap_cols) + theme(legend.text = element_text(size = 9, colour = "black"), legend.key.size = unit(10, "pt")) + guides( color = guide_legend(ncol = 1), color = guide_legend(override.aes = list(size = 10))) + ggtitle("SSBP scRNA") + NoAxes()
dim_plot <- DimPlot(pbmc.combined, split.by = "orig.ident", reduction = "umap", cols = umap_cols) + NoLegend()
pbmc.combined$orig.ident<- factor(x = pbmc.combined$orig.ident, levels = c("SS-SL", "SS-SD", "SR-SL", "SR-SD"))
dim_plot

DefaultAssay(pbmc.combined) <- "ATAC"
dim_plot <- DimPlot(pbmc.combined, split.by = "orig.ident", reduction = "umap", cols = umap_cols) 
dim_plot
```

#Linking Peaks to Genes - NLRP3 Coverage Plot

```{r, fig.asp=1, fig.width = '800px'}
pbmc.combined <- readRDS("../ATACseq SSBP PBMC/final_annotated_pbmc_combined.rds")

DefaultAssay(pbmc.combined) <- "ATAC"
library(BSgenome.Hsapiens.UCSC.hg38)
#first compute the GC content for each peak
pbmc.combined <- RegionStats(pbmc.combined, genome = BSgenome.Hsapiens.UCSC.hg38)
pbmc.combined$celltype.stim <- paste(pbmc.combined$predicted.celltype.l2, pbmc.combined$orig.ident, sep = "_")
Idents(pbmc.combined) <- "celltype.stim"
#link peaks to genes, find peaks that are significantly correlated with the expression of nearby genes

pbmc.combined <- LinkPeaks(
  object = pbmc.combined,
  peak.assay = "ATAC",
  expression.assay = "RNA",
  genes.use = c("NLRP3")
)

cov_plot <- CoveragePlot(
  object = pbmc.combined,
  region = "NLRP3",
  features = "NLRP3",
  expression.assay = "RNA",
  idents = c( "CD14 Mono_SR-SL", "CD14 Mono_SR-SD","CD14 Mono_SS-SL", "CD14 Mono_SS-SD"),
  extend.upstream = 500,
  extend.downstream = 10000,
  split.by = "orig.ident"
)



cov_plot
```



#subet monocytes - analyze NLRP3
```{r}
#subset monocytes
mono_subset <- subset(
  pbmc.combined, 
  subset = predicted.celltype.l2 %in% c("CD14 Mono", "CD16 Mono"))

cd14 <- subset(
  mono_subset,
  subset = celltype.stim %in% c(
    "CD14 Mono_SS-SL", "CD14 Mono_SS-SD",
    "CD14 Mono_SR-SL", "CD14 Mono_SR-SD"
  )
)

#extract metadata and group assignment
mono_subset$celltype.stim <- gsub("CD14 Mono", "CD16 Tcells",
                                         gsub("CD8 TEM", "CD8 Tcells",
                                              tcell_subset$celltype.stim))
cd14_df <- cd14@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  mutate(
    Group = celltype.stim,
    SaltSensitivity = case_when(
      grepl("SS-", Group) ~ "SS",
      grepl("SR-", Group) ~ "SR",
      TRUE ~ NA_character_
    ),
    Condition = case_when(
      grepl("-SL$", Group) ~ "SL",
      grepl("-SD$", Group) ~ "SD",
      TRUE ~ NA_character_
    ),
    SubjectID = paste(SaltSensitivity, Condition, sep = "_")
  )

cd14$SaltSensitivity <- cd14_df$SaltSensitivity
Idents(cd14) <- "SaltSensitivity"

##Subcluster monocytes and look at differential NLRP3 expression
#Perform subclustering
# Normalize and find variable features (if not already done)
DefaultAssay(cd14) <- "RNA"
cd14 <- NormalizeData(cd14)
cd14 <- FindVariableFeatures(cd14)
cd14 <- ScaleData(cd14)

# PCA and clustering
cd14 <- RunPCA(cd14)
cd14 <- FindNeighbors(cd14, dims = 1:10)
cd14 <- FindClusters(cd14, resolution = 0.3)  # Try a few values (e.g., 0.2–0.6)
cd14 <- RunUMAP(cd14, dims = 1:10)

# Plot UMAP with monocyte subclusters
DimPlot(cd14, group.by = "seurat_clusters", reduction = "umap", label = TRUE, pt.size = .5, label.size = 8) +
  scale_color_brewer(palette = "Set3") +
  theme_minimal(base_size = 14) + 
  ggtitle("CD14+ Monocyte Subclusters") +
  theme(axis.text.x = element_text(size = 12, hjust = 1),
        plot.title = element_text(face = "bold", size = 16),
        strip.text = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 16, face = "bold"),
        ) + theme(
  axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
  strip.text = element_text(face = "bold", size = 16),
  plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
  panel.spacing = unit(1, "lines"))


##Clusters still seaprated by treatment, trying SCTransform
# Split by group for individual SCTransform normalization
cd14_list <- SplitObject(cd14, split.by = "orig.ident")
#run SCTransform
cd14_list <- lapply(cd14_list, function(x) {
  SCTransform(x, verbose = FALSE)
})
# Select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = cd14_list, nfeatures = 3000)

# Run PrepSCTIntegration if you plan to use SCT-based integration
cd14_list <- PrepSCTIntegration(object.list = cd14_list, anchor.features = features)

# Merge all datasets back into one object
cd14 <- merge(cd14_list[[1]], y = cd14_list[-1])

# Run PCA
cd14 <- RunPCA(cd14, features = features)

library(harmony)
library(ggplot2)
# Run Harmony (integrate on condition ID or salt group)
cd14 <- RunHarmony(cd14, group.by.vars = "orig.ident", assay.use = "SCT")

# Embed into UMAP using Harmony embeddings
cd14 <- RunUMAP(cd14, reduction = "harmony", dims = 1:20)
cd14 <- FindNeighbors(cd14, reduction = "harmony", dims = 1:20)
cd14 <- FindClusters(cd14, resolution = 0.5)  # You can adjust resolution


DimPlot(cd14, group.by = "seurat_clusters", reduction = "umap", label = TRUE, label.size = 8) +
  labs(title = "CD14+ Monocyte Subclusters - Harmony + SCTransform") + 
scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 16) + 
  ggtitle("CD14+ Monocyte Subclusters-Singlets") + 
  theme(axis.text.x = element_text(size = 16, hjust = 1, angle = 45),
        axis.text.y = element_text(size = 16, hjust = 1),
        plot.title = element_text(face = "bold"),
        legend.text = element_text(size = 16),
  ) 

# Create a new metadata column for salt group (SS-SL, etc.)
cd14$SaltGroup <- cd14$celltype.stim  # assuming this exists

# Re-fetch the data from the Seurat object to ensure all metadata is included
expr_df_singlet <- FetchData(cd14, vars = c("rna_NLRP3", "orig.ident", "seurat_clusters"))

# Convert to proper factor with desired order
expr_df_singlet$SaltGroup <- factor(expr_df_singlet$orig.ident, 
                                    levels = c("SS-SL", "SS-SD", "SR-SL", "SR-SD"))

# Double-check levels (optional)
levels(expr_df_singlet$SaltGroup)

# Violin plot with jitter per cluster and salt group
ggplot(expr_df_singlet, aes(x = SaltGroup, y = rna_NLRP3, fill = SaltGroup)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.4) +
  facet_wrap(~ seurat_clusters, ncol = 3, labeller = labeller(seurat_clusters = function(x) paste("Cluster:", x))) +
theme_minimal(base_size = 14) +
  labs(
    title = "NLRP3 Expression by Monocyte Subcluster",
    x = "Salt Group",
    y = "Log-Normalized NLRP3 Expression"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 14),
    strip.text = element_text (size = 14, face = "bold")
  )

```










